Tail associations and their impact on extinction risk

### 2019-10-30 MainText.Rmd
further comments on extrisk methods

<!--FIG : pedagogical fig showing 4 types of tail associations: 
extreme left, moderate left, moderate right, extreme right-->
\begin{figure*}[!h]
\begin{center}
\includegraphics[width=15 cm]{./Results/introfig_copula.pdf}
\caption{Two stochastic quantities can have same Spearman 
correlation, but different degrees of tail association: 
(A) extreme left-tail association, (B) moderate left-tail association, (C) 
moderate right-tail association, and (D) extreme right-tail association.
In all cases, Spearman's $\rho$ was the same, $\rho =$ `r srho_common`,
up to sampling variance (the sample correlation for each dataset is displayed
on its panel). The extreme left-tail (respectively, right-tail) dependent 
case is perfectly correlated below (respectively, above) the threshold of 0.
See Methods for information on how bivariate noise was generated with
asymmetric tail associations.
\label{fig_intro_cop}}
\end{center}
\end{figure*}
<!--***DAN: Shyamolina, you put a negative sign in front of the Spearman correlations
on the C and D panels, but that is erroneous. The Spearman correlations are positive
in all four cases! Please fix!
***Shya: Yes, you are right, I fixed that.-->

@@

The variables $u(t)$ and $v(t)$ are visibly asymmetrically tail associated (Fig. \ref{fig_intro_cop}),
but we have instead used the transformed variables $s_e(t)$ and $\tilde{f}_a(t)$
in our model. In what sense are $s_e(t)$ and $\tilde{f}_a(t)$ asymmetrically
tail associated? Given two boundary quantities $l_b$ and $u_b$ such
that $0 \leq l_b < u_b \leq 1$, and letting $x(t)$ be the rank of 
$u(t)$ in the set $\{u(t) | t=1,\ldots,T\}$ and letting 
$y(t)$ be the rank of 
$v(t)$ in the set $\{v(t) | i=1,\ldots,T\}$ (smallest elements
are here defined to have rank 1),
consider the portions of the 
lines $x+y = 2l_b$ and $x+y=2u_b$ which lie within the unit square
(Fig. \ref{PartialSpearmanFig}). 
<!--***DAN: Shyamolina, please make this figure. You can start from the code that 
make the similar figure in BIVAN (Fig. 7), but change it slightly so the figure is not
*exactly* the same. E.g., you could just switch from Clayton to survival Clayton data,
and maybe slightly change the locations of the bands. The caption can also be similar,
but obviously should not mention D^2 or P, and wording should be changed slightly.
The caption will also have to say something like "This figure was reproduced largely 
following Fig. 7 of Ghosh et al".
***Shya: see below.-->

<!--FIG : schematic diagram for partial spearman correlation, adapted from BIVAN-->
\begin{figure*}[!h]
\begin{center}
\includegraphics[width=6 cm]{./Results/PartialSpearmanFig.pdf}
\caption{The partial Spearman correlation, $\cor_{lb,ub}(u, v)$, within a band can be computed
for any band to describe how the strength of association between $u$ and $v$
varies from one part of the two distributions to another. Diagonal lines 
show two bands, the data in the right/upper band showing
stronger association than those in the left/lower band. This figure was reproduced
largely following $Fig. 7$ of $Ghosh et al.(2019)$.
\label{PartialSpearmanFig}}
\end{center}
\end{figure*}

We can then define the *partial Spearman correlation*,
\begin{equation}
\cor_{l_b,u_b}(u,v) = \frac{\sum (x(t)-\mean(x))(y(t)-\mean(y))}{(n-1)\sqrt{\var(x) \var(y)}},\label{PartialSpearmanEq}
\end{equation}
\noindent where sample means and variances in this equation
are computed using all the data but the sum is computed only
over the $t$ such that $x(t)+y(t) > 2l_b$ and $x(t)+y(t)<2u_b$. 
The partial Spearman correlation is the component of the Spearman
correlation of $u$ and $v$ which can be attributed to the points
for which the ranks, $x$ and $y$, lie between the boundary lines
we defined in
the unit square (Fig. \ref{PartialSpearmanFig}). 
The quantity $\cor_{0,0.5}(u,v)-\cor_{0.5,1}(u,v)$ 
is a way to measure asymmetry of tail associations between
$u$ and $v$. And this quantity is the same as 
$\cor_{0,0.5}(s_e,\tilde{f}_a)-\cor_{0.5,1}(s_e,\tilde{f}_a)$, 
because the partial Spearman correlation, like the Spearman
correlation itself, is based on ranks. Ranks are the same
for $u(t)$ and for $s_e(t)$ because these quantities are 
related by an increasing monotonic transformation; and likewise for 
$v(t)$ and $\tilde{f}_a(t)$. Thus, in the sense captured
by the statistic $\cor_{0,0.5}-\cor_{0.5,1}$, $s_e(t)$
and $\tilde{f}_a(t)$ have the same asymmetry of tail
association as the values $u(t)$ and $v(t)$ from which they
were constructed. Rank-based approaches to understanding the nature
of association between variables have been recommended, as such
approaches are not influenced by the marginal distributions of
the quantities (e.g., @Genest2007). Marginal distributions contain no information
about association between variables, as each marginal pertains to
only one of the variables.

@@

To calculate extinction risk for a given metapopulation model, given parameters,
and a given choice of tail association,
we simulated the model for $25$ time steps, $10000$ times, using $n=5$ habitat patches.
The equilibrium value of the 
one-patch, deterministic version of the model, $P_e$ (Table \ref{tab_model_summary})
was used as the initial condition in all patches and simulations.
After each time step, if $P_i(t)$ was less than $P_e/10$ then it was set to $0$.
The whole metapopulation was considered to have gone extinct if all patches were $0$
a time step $25$.
<!--***DAN: Shyamolina, please check the above two sentences, is this what you did?
***Shya: I agreed with you upto line 482, the last line "The whole metapopulation was considered to have gone extinct if all patches were $0$ a time step $25$" I could not understand, did you write that conceptually? (then it's ok), 
or you meant I apply that in my code?
Similarly, the below line explains lines 149-150 of mydispersal.R, I am not fluent in using apply function on two dims at once e.g. c(1,3), you wrote that piece of code, I want to explain you 
on Friday gow I interpret that portion, then I can tell you accurately what I did.-->
Extinction risk was the fraction of 10000 metapopulations that went extinct. 
Estimates of extinction risk were plotted together with 95\% confidence intervals
based on a binomial distribution.

>>>>

<!--FIG : pedagogical fig showing 4 types of tail associations: 
extreme left, moderate left, moderate right, extreme right-->
\begin{figure*}[!h]
\begin{center}
\includegraphics[width=15 cm]{./Results/introfig_copula.pdf}
\caption{Two stochastic quantities can have same Spearman 
correlation, but different degrees of tail association: 
(A) extreme left-tail association, (B) moderate left-tail association, (C) 
moderate right-tail association, and (D) extreme right-tail association.
In all cases, Spearman's $\rho$ was the same, $\rho =$ `r srho_common`,
up to sampling variance (the sample correlation for each dataset is displayed
on its panel). The extreme left-tail (respectively, right-tail) dependent 
case is perfectly correlated below (respectively, above) the threshold of 0.
See Methods for information on how bivariate noise was generated with
asymmetric tail associations.
\label{fig_intro_cop}}
\end{center}
\end{figure*}

@@

The variables $u(t)$ and $v(t)$ are visibly asymmetrically tail associated (Fig. \ref{fig_intro_cop}),
but we have instead used the transformed variables $s_e(t)$ and $\tilde{f}_a(t)$
in our model. In what sense are $s_e(t)$ and $\tilde{f}_a(t)$ asymmetrically
tail associated? Given two boundary quantities $l_b$ and $u_b$ such
that $0 \leq l_b < u_b \leq 1$, and letting $x(t)$ be the rank of 
$u(t)$ in the set $\{u(t) | t=1,\ldots,T\}$ and letting 
$y(t)$ be the rank of 
$v(t)$ in the set $\{v(t) | i=1,\ldots,T\}$ (smallest elements
are here defined to have rank 1),
consider the portions of the 
lines $x+y = 2l_b$ and $x+y=2u_b$ which lie within the unit square
(Fig. \ref{PartialSpearmanFig}). 
We can then define the *partial Spearman correlation*,
\begin{equation}
\cor_{l_b,u_b}(u,v) = \frac{\sum (x(t)-\mean(x))(y(t)-\mean(y))}{(n-1)\sqrt{\var(x) \var(y)}},\label{PartialSpearmanEq}
\end{equation}
\noindent where sample means and variances in this equation
are computed using all the data but the sum is computed only
over the $t$ such that $x(t)+y(t) > 2l_b$ and $x(t)+y(t)<2u_b$. 
The partial Spearman correlation is the component of the Spearman
correlation of $u$ and $v$ which can be attributed to the points
for which the ranks, $x$ and $y$, lie between the boundary lines
we defined in
the unit square (Fig. \ref{PartialSpearmanFig}). 
The quantity $\cor_{0,0.5}(u,v)-\cor_{0.5,1}(u,v)$ 
is a way to measure asymmetry of tail associations between
$u$ and $v$. And this quantity is the same as 
$\cor_{0,0.5}(s_e,\tilde{f}_a)-\cor_{0.5,1}(s_e,\tilde{f}_a)$, 
because the partial Spearman correlation, like the Spearman
correlation itself, is based on ranks. Ranks are the same
for $u(t)$ and for $s_e(t)$ because these quantities are 
related by an increasing monotonic transformation; and likewise for 
$v(t)$ and $\tilde{f}_a(t)$. Thus, in the sense captured
by the statistic $\cor_{0,0.5}-\cor_{0.5,1}$, $s_e(t)$
and $\tilde{f}_a(t)$ have the same asymmetry of tail
association as the values $u(t)$ and $v(t)$ from which they
were constructed. Rank-based approaches to understanding the nature
of association between variables have been recommended, as such
approaches are not influenced by the marginal distributions of
the quantities (e.g., @Genest2007). Marginal distributions contain no information
about association between variables, as each marginal pertains to
only one of the variables.

<!--FIG : schematic diagram for partial spearman correlation, adapted from BIVAN-->
\begin{figure*}[!h]
\begin{center}
\includegraphics[width=6 cm]{./Results/PartialSpearmanFig.pdf}
\caption{The partial Spearman correlation, $\cor_{l_b,u_b}(u, v)$, within a band can be computed
for any band to describe how the strength of association between $u$ and $v$
varies from one part of the two distributions to another. Diagonal lines 
show two bands, the data in the right/upper band showing
stronger association than those in the left/lower band. This figure was reproduced
largely following Fig. 7 of Ghosh \emph{et al.} (2019).
\label{PartialSpearmanFig}}
\end{center}
\end{figure*}
<!--***DAN: Shyamolina, please change the u and v axis labels on the figure
to x and y, respectively, since we defined those as the ranks in the main text.-->

@@

To calculate extinction risk for a given metapopulation model, given parameters,
and a given choice of tail association,
we simulated the model for $25$ time steps, $10000$ times, using $n=5$ habitat patches.
The equilibrium value of the 
one-patch, deterministic version of the model, $P_e$ (Table \ref{tab_model_summary})
was used as the initial condition in all patches and simulations.
After each time step, if $P_i(t)$ was less than $P_e/10$ then it was set to $0$.
The whole metapopulation was considered to have gone extinct if all patches were $0$
at time step $25$.
<!--***DAN: Shyamolina, please check the above two sentences, is this what you did?
***Shya: I agreed with you upto line 482, the last line "The whole metapopulation was considered to have gone extinct if all patches were $0$ a time step $25$" I could not understand, did you write that conceptually? (then it's ok), 
or you meant I apply that in my code?
***DAN: I wrote it guessing that that was what you did. If that is not what you did,
when we should discuss, preferably as soon as possible, before Friday.
Similarly, the below line explains lines 149-150 of mydispersal.R, I am not fluent in using apply function on two dims at once e.g. c(1,3), you wrote that piece of code, I want to explain you 
on Friday gow I interpret that portion, then I can tell you accurately what I did.
***DAN: I added some comments above the function extrisk in mydispersal.R, after reading those
can you tell me by adding another comment what you did and if what I wrote here was correct?
-->
Extinction risk was the fraction of 10000 metapopulations that went extinct. 
Estimates of extinction risk were plotted together with 95\% confidence intervals
based on a binomial distribution.
